# Reproducible Research Week 2 Assignment

## Loading and preprocessing the data
```{r loaddata}
#Set working directory
setwd("C://Users/kris/Documents/GitHub/RepResearchWk2")

filename <- "data.zip"

# Checking if archive already exists locally, downloads if not present.
if (!file.exists(filename)){
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileURL, filename, method="curl")
}  

# Checking if file exists locally, unzipping if not
if (!file.exists("activity.csv")) { 
  unzip(filename) 
}

# Read Data
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r}
library(ggplot2)
total.steps <- tapply(data$steps, data$date, FUN=sum, na.rm=TRUE)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps, na.rm=TRUE)
median(total.steps, na.rm=TRUE)
```

From the data, we observe the mean is 9354.23 and the median is 10395.

## What is the average daily activity pattern?
```{r}
library(ggplot2)
averages <- aggregate(x=list(steps=data$steps), by=list(interval=data$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=averages, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken")
```

We can examine the maximum number of average steps across all days.

```{r}
averages[which.max(averages$steps),]
```

## Inputing missing values

Missing values may introduce a level of bias in our calculations.  So we can adjust for this by substituting averages in their place.  Personally, I think this is just bias of another form, but we'll go with it for the sake of learning.

```{r how_many_missing}
missing <- is.na(data$steps)
# How many missing
table(missing)
```

The following bit will replace missing fields with averages.

```{r}
# Replace each missing value with the mean value of its 5-minute interval
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled.data <- data
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```

Now, using the adjusted data set, we can make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps.

```{r}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN=sum)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps)
median(total.steps)
```

Mean and median values are higher after inputing missing data. The reason is
that in the original data, there are some days with `steps` values `NA` for 
any `interval`. The total number of steps taken in such days are set to 0s by
default. This is an a tad unfortunate, however, after replacing missing `steps` 
values with the mean `steps` of associated `interval` value, these 0 values are 
removed from the histogram of total number of steps taken each day.  Hooray!

## Are there differences in activity patterns between weekdays and weekends?
First, let's find the day of the week for each measurement in the dataset. In
this part, we use the dataset with the filled-in values.

```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN=weekday.or.weekend)
```

Now, let's make a panel plot containing plots of average number of steps taken
on weekdays and weekends.

```{r}
averages <- aggregate(steps ~ interval + day, data=filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
```